Incremental Forward Reasoning for White-Box~Proof~Search

### 2025-02-24 paper.tex
Clarify terminology around variables in intro

\newcommand{\Lam}[2]{\ensuremath{\lambda\, #1,\; #2}}
\newcommand{\All}[2]{\ensuremath{\forall\, #1,\; #2}}
\newcommand{\mvar}[1]{\ensuremath{{?#1}}}
\newcommand{\Prop}{\ensuremath{\mathrm{Prop}}}
\newcommand{\vars}{\ensuremath{\mathrm{vars}}}
\newcommand{\dom}{\ensuremath{\mathrm{dom}}}
\newcommand{\sub}{\ensuremath{\mathrm{sub}}}
\newcommand{\lvl}{\ensuremath{\mathrm{lvl}}}
\newcommand{\Rules}{\ensuremath{\mathcal{R}}}
\newcommand{\Hyps}{\ensuremath{\mathcal{H}}}
\newcommand{\States}{\ensuremath{\mathcal{S}}}
\newcommand{\CMatches}{\ensuremath{\mathcal{C}}}
\newcommand{\addHyp}{\ensuremath{\mathrm{addHyp}}}
\newcommand{\delHyp}{\ensuremath{\mathrm{delHyp}}}
\newcommand{\addMatch}{\ensuremath{\mathrm{addMatch}}}
\newcommand{\powerset}{\ensuremath{\mathcal{P}}}

@@

We first introduce some terminology for matches and for forward rules (which we identify with their underlying lemmas).
Our foundation in this paper is dependent type theory, but our technique can be adapted to other logics.

@@

Premises on which a later premise depends are \emph{variables}.
All other premises are \emph{input premises}.
For example, the lemma $\All{x~y~z}{x < y → y < z → x < z}$ has variables $x, y, z$ and (unnamed) input premises of types $x < y$ and $y < z$.
If a variable occurs in at least two premises, like $y$ above, it is \emph{shared}.

For a lemma with $n$ input premises, we assign to each input premise a unique number between $1$ and $n$.
This is the premise's \emph{slot}, which we often identify with the premise itself.
For now, we assume that the input premises are numbered consecutively, so in the above example slot~1 has type $x < y$ and slot~2 has type $y < z$.
The \emph{shared variables of slot $i$}, $\vars(i)$, are those variables that appear in the type of $i$ and also in at least one type of a slot $j < i$.

@@

A \textit{match} $m$ for a given rule $r$ with $n$ slots is a list containing at least one and at most $n$ hypotheses.
We interpret $m$ as a partial map from slots of $r$ to hypotheses, so $\dom(m) = \{i \mid 1 ≤ i ≤ |m| \}$ and for each slot $i ∈ \dom(m)$, $m(i)$ is the hypothesis assigned to $i$.
Each match $m$ must satisfy the following requirements:
\begin{enumerate}
  \item\label{cond:match:matches} For each slot $i ∈ \dom(m)$, the hypothesis $m(i)$ matches $i$.
  %\smallspace

@@

\para{Match Equivalence.}
Naively, two matches would be considered equal if they contain the same hypotheses (up to definitional equality, again implemented via RPINF).
However, consider the rule $r : \All{x~y}{P~x~y → Q~x~z → R~x}$ and the hypotheses $h₁ : P~a~b$, $h₂ : P~a~c$, $h₃ : Q~a~d$ and $h₄ : Q~a~e$.
The complete matches $m₁ ≔ [h₁, h₃]$ and $m₁' ≔ [h₁, h₄]$ contain different hypotheses, but the difference does not matter because the applications $r~h₁~h₃$ and $r~h₁~h₄$ both have type $R~a$.
Generating both $m₁$ and $m₁'$ would therefore be redundant.
Similarly, the partial matches $m₂ ≔ [h₁]$ and $m₂' ≔ [h₂]$ are redundant because the choice between $h₁$ and $h₂$ affects neither the type of any eventual application derived from $m₂$ and $m₂'$ nor the hypotheses that can be used to complete them.
Hence we should store only one of $m₂$ and $m₂'$ in the rule state of $r$.
Based on these observations, we consider two matches $m$ and $m'$ of a rule $r$ with conclusion $A$ equal if $\lvl(m) = \lvl(m')$ and for each variable $x$ that appears either in the conclusion of $r$ or in the type of a slot $i$ of $r$ with $i > \lvl(m)$ we have $\sub(m)(x) = \sub(m')(x)$.

>>>>

\newcommand{\Lam}[2]{\ensuremath{\lambda\, #1,\; #2}}
\newcommand{\All}[2]{\ensuremath{\forall\, #1,\; #2}}
\newcommand{\mvar}[1]{\ensuremath{{?#1}}}
\newcommand{\Prop}{\ensuremath{\mathrm{Prop}}}
\newcommand{\vars}{\ensuremath{\mathrm{svars}}}
\newcommand{\dom}{\ensuremath{\mathrm{dom}}}
\newcommand{\sub}{\ensuremath{\mathrm{sub}}}
\newcommand{\lvl}{\ensuremath{\mathrm{lvl}}}
\newcommand{\Rules}{\ensuremath{\mathcal{R}}}
\newcommand{\Hyps}{\ensuremath{\mathcal{H}}}
\newcommand{\States}{\ensuremath{\mathcal{S}}}
\newcommand{\CMatches}{\ensuremath{\mathcal{C}}}
\newcommand{\addHyp}{\ensuremath{\mathrm{addHyp}}}
\newcommand{\delHyp}{\ensuremath{\mathrm{delHyp}}}
\newcommand{\addMatch}{\ensuremath{\mathrm{addMatch}}}
\newcommand{\powerset}{\ensuremath{\mathcal{P}}}

@@

We first introduce some terminology for matches and for forward rules (which we identify with their underlying lemmas).
We use dependent type theory as a foundation, but our technique generalises to other logics.

@@

Using somewhat non-standard terminology, we call premises on which a later premise depends \emph{variables}.
These are typically terms of non-propositional types, e.g.\ natural numbers or lists.
The non-variable premises, whose types are typically propositions, are called \emph{slots}.
For example, the lemma $\All{(x~y~z : ℕ)}{x < y → y < z → x < z}$ has variables $x, y, z$ and slots of types $x < y$ and $y < z$.
If more than two premises depend on a variable (e.g.\ $y$), the variable is \emph{shared}.
Note that premises on which only the conclusion depends are not considered variables.

If a lemma has $n$ slots, each slot is assigned a unique number between $1$ and $n$, and we identify slots with their numbers.
For now, we assume that the slots are numbered consecutively, so in the example above slot~1 has type $x < y$ and slot~2 has type $y < z$.
The \emph{shared variables of slot $i$}, $\vars(i)$, are those variables that appear in the type of $i$ and also in at least one type of a slot $j < i$.

@@

A \textit{match} $m$ for a given rule $r$ with $n$ slots is a list containing at least one and at most $n$ hypotheses.
We interpret $m$ as a partial map from slots of $r$ to hypotheses, so $\dom(m) = \{i \mid 1 ≤ i ≤ |m| \}$ and for each slot $i ∈ \dom(m)$, $m(i)$ is the hypothesis assigned to $i$.
A match $m$ must additionally satisfy the following requirements:
\begin{enumerate}
  \item\label{cond:match:matches} For each slot $i ∈ \dom(m)$, the hypothesis $m(i)$ matches $i$.

@@

\para{Match Equivalence.}
Naively, two matches would be considered equal if they contain the same hypotheses (up to definitional equality, again implemented via RPINF).
However, consider the rule $r : \All{x~y}{P~x~y → Q~x~z → R~x}$ and the hypotheses $h₁ : P~a~b$, $h₂ : P~a~c$, $h₃ : Q~a~d$ and $h₄ : Q~a~e$.
The complete matches $m₁ ≔ [h₁, h₃]$ and $m₁' ≔ [h₁, h₄]$ contain different hypotheses, but the difference does not matter because the applications $r~h₁~h₃$ and $r~h₁~h₄$ both have type $R~a$.
Generating both $m₁$ and $m₁'$ would therefore be redundant.
Similarly, the partial matches $m₂ ≔ [h₁]$ and $m₂' ≔ [h₂]$ are redundant because the choice between $h₁$ and $h₂$ affects neither the type of any eventual application derived from $m₂$ and $m₂'$ nor the hypotheses that can be used to complete them.
Hence we should store only one of $m₂$ and $m₂'$ in the rule state of $r$.
Based on these observations, we consider two matches $m$ and $m'$ of a rule $r$ with conclusion $A$ equal if $\lvl(m) = \lvl(m')$ and for each premise $x$ that appears either in the conclusion of $r$ or in the type of a slot $i$ of $r$ with $i > \lvl(m)$ we have $\sub(m)(x) = \sub(m')(x)$.
